\textit{Supplementary materials for:} The psychological arrow of time drives temporal asymmetries in inferring unobserved past and future events

### 2023-09-22 paper/main.tex
inputting data + details from meta analysis
(might want to put some of the tables in the supplement instead of the main text)
finished adding meta analysis to main text, also added meta analysis tables to supplemental materials

% tables
\newcommand{\stimDescription}{S1}
\newcommand{\metaTable}{S2}

@@

\caption{\textbf{Meta analysis.} We used natural language processing to
automatically identify references to past or future events across a variety of
sources. \textbf{A. Numbers of past and future events in \textit{The Chair},
Season 1, Episodes 1--6.} The bar heights indicate the raw numbers of manually
identified (lighter shading) and automatically identified (darker shading) past
and future events from each episode (color). We used Episode 1 from this series
as the stimulus in our replication experiment. \textbf{B. Proportions of past
and future events in \textit{The Chair}, Season 1, Episodes 1--6.} The Panel is
in the same format as Panel A, but here the bar heights have been divided by
the total numbers of past and future events (per episode). \textbf{C.
Proportions of past and future events in movies, television shows, and natural
conversations.} As in Panel B, the bar heights denote the proportions of past
and future events detected in each dataset (color). The datasets are described
in Table~\metaTable. Error bars denote bootstrap-estimated 95\% confidence
intervals.}

@@

The above analyses show that characters in the television shows we used as
stimuli in our main experiment and replication experiment refer more often to
the past than to the future. This appears to bias participants' inferences
about the past and future. But how universal is this pattern? For example, were
the television shows we happened to select for our experiment representative of
television shows more generally? Or perhaps media created for entertainment
purposes tends to have a bias towards the past in order to keep the story
engaging and unpredictable. To better understand temporal biases in
conversations, we carried out a meta analysis using extracted conversation data
from several large datasets, comprising a total of over 440 million
conversations from over 17 million documents. The data comprised transcripts
from television shows and popular films, novels, and spoken and written
utterances from natural conversations. A summary of the data we analyzed may be
found in Table~\metaTable. As summarized in Figure~\ref{fig:meta-analysis}, we
used natural language processing to identify references to past or future
events in each conversation (also see \textit{Meta analysis of conversation
data}).

@@

In all, across all of the datasets we examined in our meta analysis, we
identified a total of 36,008,500 references to past or future events. A total
of 19,464,741 (54.06\%) of these were references to past events, and the
remaining 16,543,759 (45.94\%) were references to future events. We also
computed the average proportions of references to past and future events across
documents within each individual dataset. Across the 12 datasets we examined
(Fig.~\ref{fig:meta-analysis}, Tab.~\metaTable), there were significantly more
references to the past than the to the future (mean $\pm$ standard deviation:
$58.99\% \pm 7.28\%$; $t(11) = 4.28, p = 0.0013$). This bias towards the past
also held for each dataset individually ($t$s $\geq 5.14$, $p$s $< 0.01$)
except for one dataset, ``Persuasion for Good,'' which comprised natural
conversations between pairs of Amazon Mechanical Turk workers wherein one
participant tried to convince the other participant to donate to a charity in
the future. In that dataset, references to the future were significantly more
common than references to the past ($t(11438) = -22.65, p < 0.001$). This
latter example provided a nice sanity check for verifying that our general
approach was not itself biased in favor of the past, e.g., even in
conversations that were actually biased towards the future. Taken together, the
results from our meta analysis indicate that people tend to refer to the past
more than they refer to the future, across a wide variety of situations
(including in both fictional and real conversations). Although (as in the
Persuasion for Good dataset) there may be specific exceptions to this bias, it
seems that a bias in favor of the past is a common element of many (and perhaps
even \textit{most}) human conversations.

@@

>>>>

% tables
\newcommand{\stimDescription}{S1}
\newcommand{\regExpTable}{S2}
\newcommand{\posTags}{S3}
\newcommand{\pastKeys}{S4}
\newcommand{\futureKeys}{S5}
\newcommand{\metaAnalysisDatasets}{S6}

@@

\caption{\textbf{Meta analysis.} We used natural language processing to
automatically identify references to past or future events across a variety of
sources. \textbf{A. Numbers of past and future events in \textit{The Chair},
Season 1, Episodes 1--6.} The bar heights indicate the raw numbers of manually
identified (lighter shading) and automatically identified (darker shading) past
and future events from each episode (color). We used Episode 1 from this series
as the stimulus in our replication experiment. \textbf{B. Proportions of past
and future events in \textit{The Chair}, Season 1, Episodes 1--6.} The Panel is
in the same format as Panel A, but here the bar heights have been divided by
the total numbers of past and future events (per episode). \textbf{C.
Proportions of past and future events in movies, television shows, and natural
conversations.} As in Panel B, the bar heights denote the proportions of past
and future events detected in each dataset (color). The datasets are described
in Table~\metaAnalysisDatasets. Error bars denote bootstrap-estimated 95\% confidence
intervals.}

@@

The above analyses show that characters in the television shows we used as
stimuli in our main experiment and replication experiment refer more often to
the past than to the future. This appears to bias participants' inferences
about the past and future. But how universal is this pattern? For example, were
the television shows we happened to select for our experiment representative of
television shows more generally? Or perhaps media created for entertainment
purposes tends to have a bias towards the past in order to keep the story
engaging and unpredictable. To better understand temporal biases in
conversations, we carried out a meta analysis using extracted conversation data
from several large datasets, comprising a total of over 440 million
conversations from over 17 million documents. The data comprised transcripts
from television shows and popular films, novels, and spoken and written
utterances from natural conversations. A summary of the data we analyzed may be
found in Table~\metaAnalysisDatasets. As summarized in Figure~\ref{fig:meta-analysis}, we
used natural language processing to identify references to past or future
events in each conversation (also see \textit{Meta analysis of conversation
data}).

@@

In all, across all of the datasets we examined in our meta analysis, we
identified a total of 36,008,500 references to past or future events. A total
of 19,464,741 (54.06\%) of these were references to past events, and the
remaining 16,543,759 (45.94\%) were references to future events. We also
computed the average proportions of references to past and future events across
documents within each individual dataset. Across the 12 datasets we examined
(Fig.~\ref{fig:meta-analysis}, Tab.~\metaAnalysisDatasets), there were significantly more
references to the past than the to the future (mean $\pm$ standard deviation:
$58.99\% \pm 7.28\%$; $t(11) = 4.28, p = 0.0013$). This bias towards the past
also held for each dataset individually ($t$s $\geq 5.14$, $p$s $< 0.01$)
except for one dataset, ``Persuasion for Good,'' which comprised natural
conversations between pairs of Amazon Mechanical Turk workers wherein one
participant tried to convince the other participant to donate to a charity in
the future. In that dataset, references to the future were significantly more
common than references to the past ($t(11438) = -22.65, p < 0.001$). This
latter example provided a nice sanity check for verifying that our general
approach was not itself biased in favor of the past, e.g., even in
conversations that were actually biased towards the future. Taken together, the
results from our meta analysis indicate that people tend to refer to the past
more than they refer to the future, across a wide variety of situations
(including in both fictional and real conversations). Although (as in the
Persuasion for Good dataset) there may be specific exceptions to this bias, it
seems that a bias in favor of the past is a common element of many (and perhaps
even \textit{most}) human conversations.

@@

\subsection*{Meta analysis}

At a high level, the goal of our meta analysis was to predict in-text
references to past and future events. Manually identifying these references is
labor and time intensive, so it is impractical to scale up manual tagging to
millions of documents. Instead, we defined a set of heuristics for
\textit{predicting} when text is referring to real or hypothetical past or
future events. Our approach comprises four main steps.

First, we use the \texttt{nltk} package~\citep{BirdEtal09} to segment each
document into individual sentences. Each sentence is processed independently of
the others. Second, we handle contractions using the \texttt{contractions} package
(e.g., ``we'll'' is split into ``we will,'' and so on).  Third, we define two
sets of ``keywords'' (words and phrases) that tend to be indicative of referring
to the past (Tab.~\pastKeys) or future (Tab.~\futureKeys). We used
ChatGPT~\citep{ChatGPT} to generate each list, with exactly 50 templates per list,
using the following prompt:

\begin{quote}
\texttt{I'm designing a heuristic algorithm for identifying references (in text) to
past and future events. Part of the algorithm will involve looking for specific
keywords or phrases that suggest that the text is referring to something that
happened (or will happen) in the past and/or future. Could you help me generate
a list of 50 keywords or phrases to include in each list (one list for
identifying references to the past and a second list for identifying references
to the future)? I'd like to be able to paste the lists you generate into two
plain text documents with one row per keyword or phrase, and no other content.
Please output the lists as a "code" block (enclosed by '''...''').}
\end{quote}
Fourth, we use part-of-speech tagging (again, using the \texttt{nltk} package) to look for
verbs or verb phrases that are in past or future tenses.  After the words were
tagged with their predicted parts of speech, we use regular expressions
(applied to the sequences of tags) to label each verb or verb phrase with a
human readable verb form (e.g., ``future perfect continuous passive,''
``conditional perfect continuous passive,'' and so on).  The regular expressions
we used to generate these labels are shown in Table~\regExpTable, and the part of
speech tags are defined in Table~\posTags.

We treated each keyword match (of past or future keywords) as a single
``reference'' (to a past or future event, respectively), and if any past or
future verb forms were detected we treat those as (up to) one additional
reference. We then tallied up the numbers of past and/or future references
across sentences within the given document. The meta analysis results reported
in Figure~\ref{fig:meta-analysis}C display the average numbers of references
aggregated across all documents within each dataset we analyzed (described in
Tab.~\metaAnalysisDatasets).

### 2023-09-22 paper/supplement.tex
finished adding meta analysis to main text, also added meta analysis tables to supplemental materials

>>>>

\begin{table}
    \footnotesize
    \caption{List of datasets analyzed in our meta-analysis (see Fig. 8).}
    \label{tab:datasets}
    \begin{tabular}{p{0.6in}|l|p{2in}|l|p{0.5in}|p{0.5in}|p{0.5in}}
        \toprule
        \textbf{Dataset} & \textbf{Short name} & \textbf{Description} & \textbf{Category} & \textbf{Number of observations} & \textbf{Ob\-ser\-va\-tion type} & \textbf{Number of words} \\
        \midrule
        Internet Movie Script Database & IMSDb & A collection of transcripts from roughly 1000 popular movies. & Film & 1091 & Transcript & 26023348 \\
        \midrule
        Movie Dialogues Dataset & Movies & A large collection of fictional conversations extracted from raw movie scripts. & Film & 304713 & Utterance & 3209921 \\
        \midrule
        Switchboard Dialog Act Corpus & Switchboard & A collection of five-minute telephone conversations between two participants, annotated with speech act tags. & Speech & 122646 & Utterance & 2052779 \\
        \midrule
        Supreme Court Corpus & SCOTUS & A collection of cases from the U.S. Supreme Court, along with transcripts of oral arguments. & Speech & 1700789 & Utterance & 71889094 \\
        \midrule
        Tennis Interviews & Tennis & Transcripts for tennis singles post-match press conferences for major tournaments between 2007 to 2015. & Speech & 163948 & Utterance & 7043118 \\
        \midrule
        Persuasion for Good Corpus & PfG & A collection of online conversations generated by Amazon Mechanical Turk workers, where one participant (the persuader) tries to convince the other (the persuadee) to donate to a charity. & Speech & 20932 & Utterance & 351759 \\
        \midrule
        Intelligence Squared Debates Corpus & IQ2 & This dataset contains transcripts of debates held as part of Intelligence Squared Debates. & Speech & 26562 & Utterance & 1898509 \\
        \midrule
        Group Affect and Performance Corpus & GAP & Group members completed a Winter Survival Task (WST), a group decision-making exercise where participants must rank 15 items according to their importance in a hypothetical plane crash scenario. Participants first rank the items individually. Then, each group was given a maximum of 15 minutes to complete the WST. The group’s conversations and deliberations during this task were recorded as conversations in this dataset. & Speech & 8009 & Utterance & 45989 \\
        \midrule
        The Chair & Chair & Scraped transcripts from The Chair, Season 1. & Television & 6 & Transcript & 19197 \\
        \midrule
        Friends Corpus & Friends & A collection of all the conversations that occurred over 10 seasons of Friends, a popular American TV sitcom that ran in the 1990s. & Television & 67373 & Utterance & 622894 \\
        \midrule
        Gutenberg Dialogue Dataset & Gutenberg & Dialogues extracted from the Project Gutenberg collection. & Text & 14773741 & Utterance & 327519461 \\
        \midrule
        Reddit Corpus & Reddit & A collection of Corpuses of Reddit data built from Pushshift.io Reddit Corpus. Each Corpus contains posts and comments from an individual subreddit from its inception until Oct 2018. & Text & 74468 & Utterance & 3080662 \\
        \bottomrule
    \end{tabular}
\end{table}

\begin{table}
    \footnotesize
    \centering

    \caption{Part of speech pattern templates used to identify verb tenses in
    our meta analysis. Part of speech tags are represented between angled
    brackets, and pipes ($|$) denote instances where \textit{any} of the two or
    more parts of speech are considered a match when they appear at the given
    position in the indicated sequence. The plus signs ($+$) denote that one or
    more repetitions of the given sequence within a single sentence are still
    counted as a ``single'' instance of the given tense. Part of speech tags
    are defined in Table~\ref{tab:pos_tags}.}

    \begin{tabular}{r|l}
    \toprule
      \textbf{Tense} & \textbf{Pattern template} \\
      \midrule
      conditional continuous & $<$MD$>$$<$BE$>$$<$VBG$|$HVG$|$BEG$>$$+$ \\
      \midrule
      conditional continuous passive & $<$MD$>$$<$BE$>$$<$BEG$>$$<$VBN$|$VBD$>$$+$ \\
      \midrule
      conditional indefinite & $<$MD$>$$<$BE$|$DO$|$VB$|$HV$>$$+$ \\
      \midrule
      conditional indefinite passive & $<$MD$>$$<$BE$>$$<$VBN$|$VBD$>$$+$ \\
      \midrule
      conditional perfect & $<$MD$>$$<$HV$>$$<$HVN$|$BEN$|$VBN$|$VBD$>$$+$ \\
      \midrule
      conditional perfect continuous & $<$MD$>$$<$HV$>$$<$BEN$>$$<$VBG$|$HVG$|$BEG$>$$+$ \\
      \midrule
      conditional perfect passive & $<$MD$>$$<$HV$>$$<$BEN$>$$<$VBN$|$VBD$>$$+$ \\
      \midrule
      future continuous & $<$MDF$>$$<$BE$>$$<$VBG$|$HVG$|$BEG$>$$+$ \\
      \midrule
      future continuous passive & $<$MDF$>$$<$BE$>$$<$BEG$>$$<$VBN$|$VBD$>$$+$ \\
      \midrule
      future indefinite & $<$MDF$>$$<$BE$|$DO$|$VB$|$HV$>$$+$ \\
      \midrule
      future indefinite passive & $<$MDF$>$$<$BE$>$$<$VBN$|$VBD$>$$+$ \\
      \midrule
      future perfect & $<$MDF$>$$<$HV$>$$<$HVN$|$BEN$|$VBN$|$VBD$>$$+$ \\
      \midrule
      future perfect continuous & $<$MDF$>$$<$HV$>$$<$BEN$>$$<$VBG$|$HVG$|$BEG$>$$+$ \\
      \midrule
      future perfect continuous passive & $<$MDF$>$$<$HV$>$$<$BEN$>$$<$BEG$>$$<$VBN$|$VBD$>$$+$ \\
      \midrule
      future perfect passive & $<$MDF$>$$<$HV$>$$<$BEN$>$$<$VBN$|$VBD$>$$+$ \\
      \midrule
      infinitive & $<$TO$>$$<$BE$|$HV$|$VB$>$$+$ \\
      \midrule
      past continuous & $<$BED$|$BEDZ$>$$<$VBG$|$HVG$|$BEG$>$$+$ \\
      \midrule
      past continuous passive & $<$BED$|$BEDZ$>$$<$BEG$>$$<$VBN$|$VBD$>$$+$ \\
      \midrule
      past indefinite & $<$DOD$>$$<$VB$|$HV$|$DO$>$$|$$<$BEDZ$|$BED$|$HVD$|$VBN$|$VBD$>$$+$ \\
      \midrule
      past indefinite passive & $<$BED$|$BEDZ$>$$<$VBN$|$VBD$>$$+$ \\
      \midrule
      past perfect & $<$HVD$>$$<$BEN$|$VBN$|$HVD$|$HVN$>$$+$ \\
      \midrule
      past perfect continuous & $<$HVD$>$$<$BEN$>$$<$HVG$|$BEG$|$VBG$>$$+$ \\
      \midrule
      past perfect continuous passive & $<$HVD$>$$<$BEN$>$$<$BEG$>$$<$VBN$|$VBD$>$$+$ \\
      \midrule
      past perfect passive & $<$HVD$>$$<$BEN$>$$<$VBN$|$VBD$>$$+$ \\
      \midrule
      present continuous & $<$BEM$|$BER$|$BEZ$>$$<$BEG$|$VBG$|$HVG$>$$+$ \\
      \midrule
      present continuous passive & $<$BEM$|$BER$|$BEZ$>$$<$BEG$>$$<$VBN$|$VBD$>$$+$ \\
      \midrule
      present indefinite & $<$DO$|$DOZ$>$$<$DO$|$HV$|$VB$>$$+$$|$$<$DO$|$HV$|$VB$|$BEZ$|$DOZ$|$BER$|$HVZ$|$BEM$|$VBZ$>$$+$ \\
      \midrule
      present indefinite passive & $<$BEM$|$BER$|$BEZ$>$$<$VBN$|$VBD$>$$+$ \\
      \midrule
      present perfect & $<$HV$|$HVZ$>$$<$BEN$|$HVD$|$VBN$|$VBD$>$$+$ \\
      \midrule
      present perfect continuous & $<$HV$|$HVZ$>$$<$BEN$>$$<$VBG$|$BEG$|$HVG$>$$+$ \\
      \midrule
      present perfect continuous passive & $<$HV$|$HVZ$>$$<$BEN$>$$<$BEG$>$$<$VBN$|$VBD$>$$+$ \\
      \midrule
      present perfect passive & $<$HV$|$HVZ$>$$<$BEN$>$$<$VBN$|$VBD$>$$+$ \\
      \bottomrule
    \end{tabular}
    \label{tab:regexp_templates}
\end{table}

\begin{table}[h!]
    \centering
    \caption{Part of speech tag definitions.}
    \begin{tabular}{r|l}
        \toprule
        \textbf{Tag} & \textbf{Part of speech} \\
        \midrule
        BE & Be \\
        \midrule
        BEG & Related to beginning or gerundive form \\
        \midrule
        BEM & Am (first person singular present of BE) \\
        \midrule
        BEN & Been (past participle of BE) \\
        \midrule
        BER & Are (second person singular and all plural present of BE) \\
        \midrule
        BEZ & Is (third person singular present of BE) \\
        \midrule
        BED & Were (past plural of BE) \\
        \midrule
        BEDZ & Was (past singular of BE) \\
        \midrule
        DO & Do \\
        \midrule
        DOD & Did (past tense of DO) \\
        \midrule
        DOZ & Do, 3rd person singular present \\
        \midrule
        HV & Have (archaic form for historical texts) \\
        \midrule
        HVD & Have, past tense (archaic) \\
        \midrule
        HVG & Have, gerund or present participle (archaic) \\
        \midrule
        HVN & Have, past participle (archaic) \\
        \midrule
        HVZ & Have, 3rd person singular present (archaic) \\
        \midrule
        MD & Modal \\
        \midrule
        MDF & Modal, future tense \\
        \midrule
        TO & To \\
        \midrule
        VB & Verb, base form \\
        \midrule
        VBD & Verb, past tense \\
        \midrule
        VBG & Verb, gerund or present participle \\
        \midrule
        VBN & Verb, past participle \\
        \midrule
        VBP & Verb, non-3rd person singular present \\
        \midrule
        VBZ & Verb, 3rd person singular present \\
        \bottomrule
    \end{tabular}
    \label{tab:pos_tags}
  \end{table}

  \begin{table}[h!]
    \centering
    \caption{Keywords and phrases used to identify references to the past.}
    \begin{tabular}{l l l l l}
    \toprule
    ago & elapsed & last month & olden days & thus far \\
    \midrule
    already & expired & last night & once & to date \\
    \midrule
    antiquity & final & last quarter & once upon a time & up to now \\
    \midrule
    back when & formerly & last season & previously & used to \\
    \midrule
    before & had & last semester & recently & used to be \\
    \midrule
    bygone & heretofore & last time & said & was \\
    \midrule
    ceased & historically & last week & since & were \\
    \midrule
    concluded & hitherto & last year & so far & wrote \\
    \midrule
    did & in the past & long ago & terminated & yesterday \\
    \midrule
    earlier & in those days & made & then & yesteryear \\
    \bottomrule
    \end{tabular}
    \label{tab:keywords_past}
  \end{table}

  \begin{table}[h!]
    \centering
    \caption{Keywords and phrases used to identify references to the future.}
    \begin{tabular}{l l l l l}
        \toprule
        after & going to & later & next time & shortly \\
        \midrule
        anticipated & hereafter & later on & next week & some day \\
        \midrule
        can & imminently & looming & next year & soon \\
        \midrule
        could & impending & may & on the horizon & subsequent \\
        \midrule
        down the line & in the cards & might & plan to & succeeding \\
        \midrule
        eventual & in the future & next month & predicted & to be \\
        \midrule
        eventually & in the works & next quarter & prospective & tomorrow \\
        \midrule
        forthcoming & in time & next season & scheduled to & upcoming \\
        \midrule
        futuristic & intend to & next semester & shall & will \\
        \bottomrule
    \end{tabular}
    \label{tab:keywords_future}
  \end{table}

  \newpage
